---
title: "rnaseq2tracks final report"
date: "`r Sys.Date()`"
params:
  outdir: ""
  config: ""
  samplesheet: ""
  species: "mouse"
  layout: "PE"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    self_contained: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)
suppressPackageStartupMessages({
  library(knitr)
  library(kableExtra)
  library(data.table)
  library(ggplot2)
})
out <- params$outdir
show_table <- function(path, ...) {
  if (!file.exists(path)) return(cat("Not available."))
  tab <- tryCatch(fread(path, ...), error=function(e) NULL)
  if (is.null(tab) || nrow(tab) == 0) return(cat("No rows."))
  kable(tab, "html", digits=3) %>%
    kable_styling(bootstrap_options=c("striped", "condensed"), full_width=FALSE) %>%
    scroll_box(width="100%")
}
embed_images <- function(paths, width="90%") {
  paths <- paths[file.exists(paths)]
  if (length(paths) == 0) return(cat("No plots available."))
  for (p in paths) {
    cat("\n\n**", basename(p), "**\n\n", sep="")
    print(include_graphics(p))
  }
}
```

## Analysis overview

This report summarizes an `rnaseq2tracks` run. Technical-library and lane rows are
processed independently through alignment, then merged into biological samples
before counting, track normalization, differential expression and enrichment.

```{r overview}
prov <- file.path(out, "metadata", "run_provenance.tsv")
if (file.exists(prov)) show_table(prov, header=FALSE, col.names=c("Field", "Value"))
```

## Configuration

```{r configuration}
cfg_vars <- c("PROJECT_ID", "SPECIES", "GENOME_ASSEMBLY", "LIBRARY_LAYOUT",
              "DESIGN_FORMULA", "RUN_FASTQSCREEN", "RUN_RSEQC", "RUN_DE",
              "RUN_ENRICHMENT", "REGULAR_CHROMS_ONLY", "CHROMOSOME_NAMING")
lines <- if (file.exists(params$config)) readLines(params$config, warn=FALSE) else character()
values <- vapply(cfg_vars, function(key) {
  hit <- grep(paste0("^", key, "="), lines, value=TRUE)
  if (length(hit)) sub("^[^=]+=?", "", hit[1]) else NA_character_
}, character(1))
kable(data.frame(Parameter=cfg_vars, Value=values), "html") %>%
  kable_styling(bootstrap_options=c("striped", "condensed"), full_width=FALSE)
```

## Samples and technical-replicate merging

```{r samples}
show_table(file.path(out, "metadata", "validated_samples.tsv"))
```

```{r merge_audit}
show_table(file.path(out, "metadata", "technical_merge_audit.tsv"))
```

The sample-level count matrix and final BAM contain the sum/union of all lane
rows assigned to that `sample_id`. Technical rows are not treated as independent
replicates in DESeq2.

## Read and alignment QC

The final MultiQC report combines raw/trimmed FastQC, FastQ Screen, STAR and
RSeQC outputs. Open `multiQC/final/multiQC_final.html` for interactive details.

```{r star_qc}
show_table(file.path(out, "07_qc", "star", "star_alignment_summary.tsv"))
```

### Library strandedness

```{r infer_experiment}
ie_dir <- file.path(out, "07_qc", "rseqc", "infer_experiment")
files <- if (dir.exists(ie_dir)) list.files(ie_dir, pattern="[.]txt$", full.names=TRUE) else character()
if (length(files)) {
  txt <- rbindlist(lapply(files, function(path) {
    x <- readLines(path, warn=FALSE)
    data.table(sample=sub("_infer_experiment[.]txt$", "", basename(path)),
               summary=paste(x[grepl("Fraction of reads", x)], collapse="; "))
  }), fill=TRUE)
  kable(txt, "html") %>% kable_styling(bootstrap_options=c("striped", "condensed"))
} else cat("infer_experiment outputs are not available.")
```

### Gene-body coverage

Gene-body coverage is an important RNA integrity and 5'/3' bias diagnostic.
Large systematic differences between samples should be reviewed before biological
interpretation.

```{r genebody, results='asis'}
gb_dir <- file.path(out, "07_qc", "rseqc", "genebody")
gb <- if (dir.exists(gb_dir)) list.files(gb_dir, pattern="[.](png|pdf)$", full.names=TRUE) else character()
embed_images(head(gb, 24))
```

## Expression and sample relationships

```{r size_factors}
show_table(file.path(out, "analysis", "counts", "size_factors.tsv"))
```

```{r global_plots, results='asis'}
fig_dir <- file.path(out, "analysis", "figures")
figs <- if (dir.exists(fig_dir)) list.files(fig_dir, pattern="[.]png$", full.names=TRUE) else character()
embed_images(figs)
```

## Differential expression

DESeq2 operates on biological samples after technical-lane count summation.
Reported fold changes are shrunken with apeglm when available (ashr fallback).

```{r de_summary}
de_dir <- file.path(out, "analysis", "DE")
sig_files <- if (dir.exists(de_dir)) list.files(de_dir, pattern="_significant[.]tsv$", full.names=TRUE) else character()
if (length(sig_files)) {
  summary <- rbindlist(lapply(sig_files, function(path) {
    tab <- tryCatch(fread(path), error=function(e) data.table())
    data.table(contrast=sub("_significant[.]tsv$", "", basename(path)),
               significant_genes=nrow(tab),
               up=sum(tab$log2FoldChange > 0, na.rm=TRUE),
               down=sum(tab$log2FoldChange < 0, na.rm=TRUE))
  }))
  kable(summary, "html") %>% kable_styling(bootstrap_options=c("striped", "condensed"))
} else cat("No significant-gene tables were produced.")
```

```{r de_plots, results='asis'}
de_plots <- if (dir.exists(de_dir)) list.files(de_dir, pattern="_(volcano_shrunk|MA_shrunk)[.]png$", full.names=TRUE) else character()
embed_images(de_plots)
```

## Functional enrichment

For every contrast, the workflow attempts ORA and ranked GSEA against GO BP/MF/CC,
KEGG, Reactome and MSigDB Hallmarks. Empty result tables are valid when no gene
set passes the configured threshold.

```{r enrichment_summary}
enr_dir <- file.path(out, "analysis", "enrichment")
enr_files <- if (dir.exists(enr_dir)) list.files(enr_dir, pattern="_(ORA|GSEA)_.+[.]tsv$", recursive=TRUE, full.names=TRUE) else character()
if (length(enr_files)) {
  enr <- rbindlist(lapply(enr_files, function(path) {
    tab <- tryCatch(fread(path), error=function(e) data.table())
    rel <- sub(paste0("^", normalizePath(enr_dir, winslash="/"), "/?"), "", normalizePath(path, winslash="/"))
    data.table(result=rel, terms=nrow(tab))
  }))
  kable(enr, "html") %>% kable_styling(bootstrap_options=c("striped", "condensed")) %>%
    scroll_box(width="100%", height="500px")
} else cat("No enrichment tables are available.")
```

## Deliverables and provenance

```{r outputs}
dirs <- c("fastQC", "fastQScreen", "multiQC", "STARlogs", "STARgeneCounts",
          "07_qc", "bams", "bedGraph", "bigwig", "analysis", "reports")
status <- vapply(dirs, function(d) if (dir.exists(file.path(out, d))) "present" else "missing", character(1))
kable(data.frame(Directory=dirs, Status=status), "html") %>%
  kable_styling(bootstrap_options=c("striped", "condensed"), full_width=FALSE)
```

The human-readable master log is `logs/rnaseq2tracks.log`; current/final machine-
readable state is `metadata/run_status.tsv`; the immutable run summary is
`metadata/run_provenance.tsv`.
